---
title: "In-class Exercise 6: Preparation for Take-home Exercise 2"
author: "PHAM Hung Son"
date: "Oct 04, 2025"
date-modified: "last-modified"
format: 
  html:
    toc-depth: 4
    toc-expand: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

**Loading R packages**

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr, rvest)
```

**Importing data**

```{r}
BusStop = st_read(dsn = "data/ltadata/",
                  layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_transform(crs = 3414)
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
write_rds(mpsz, "data/rds/mpsz.rds")
```

```{r}
odbus <- read_csv("data/ltadata/origin_destination_bus_202508.csv")
```

**Extract Bus Stops within SG**

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = FALSE)
```

```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```

**Deriving Analytical Hexagon**

```{r}
hexagon <- st_make_grid(mpsz,
                        cellsize = 375,
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()
```

**Filtering hexagons with bus stops**

```{r}
hexagon$n_collisions =
  lengths(st_intersects(
    hexagon, BusStop_in_SG))

hexagon <- filter(hexagon, n_collisions > 0)
```

**Assigning ids to each hexagon**

```{r}
hexagon <- hexagon %>%
  select(, -n_collisions) #optional can remove

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))) %>%
  as.factor()

kable(head(hexagon))
```

**Preparing Trip Generation Data**

Cleaning the data

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
kable(head(trips))
```

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))
kable(head(bs_hex))
```

Adding HEX_ID into bus trips data

```{r}
trips <- inner_join(trips, bs_hex)
kable(head(trips))
```

Aggregating TRIPS based on HEX_ID

In the code chunk below, `group_by()` and `summarise()` is used to aggregate *TRIPS* by *HEX_ID*, *DAY_TYPE* and *HOUR_OF_DAY*.

```{r}
trips <- trips %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS))
```
