---
title: "In-class Exercise 3b"
author: "Pham Hung Son"
date: "September 13, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

**Loading the R package**

```{r}
pacman::p_load(tidyverse, sf, tmap, httr)
```

**Importing ACRA data**

-   importing an ACRA data set into RStudio as tibble data frame, and

-   drop the unnecessary fields such as fields with excessive missing values or fields with only single values.

```{r}
folder_path <- "data/aspatial/ACRA"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA*.*\\.csv$",
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)
```

**Saving ACRA data**

code chunk to save the consolidated ACRA data.

```{r}
write_rds(acra_data,
          "data/rds/acra_data.rds")
```

**Tidying ACRA data**

-   select the target businesses by using SSIC code,

-   derive the year and month fields from *registration_incorporation_date* field,

-   tidy postal code values to ensure avoid 5-digit postal codes by mistake, and

-   select businesses registered in 2025. (first 6 months)

```{r}
biz_56111 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 56111) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date,
                            label = TRUE,
                            abbr = TRUE)) %>%
  mutate(postal_code = str_pad(postal_code, width = 6, side = "left", pad = "0")) %>%
  filter(YEAR == 2025, MONTH_NUM <= 6)
```

Geocoding

```{r}
postcodes <- unique(biz_56111$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )
  
  res <- GET(url, query = query)
  json <- content(res)
  
  if(json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}
```

```{r}
found <- found %>%
  select(1:10)
```

Appending the location info

```{r}
biz_56111 = biz_56111 %>%
  left_join(found,
            by = c('postal_code'= 'POSTAL'))
```

save the tidied data into rds format for future use.

```{r}
write_rds(biz_56111, "data/rds/biz_56111.rds")
```

convert biz_56111 tibble data frame into sf data frame

```{r}
biz_56111_sf <- st_as_sf(biz_56111, 
                         coords = c("X","Y"),
                         crs=3414) 
```

**Visualising the distribution**

code chunk to reveal the distribution of newly registered businesses by month.

```{r}
ggplot(data = biz_56111,
       aes(x = MONTH_ABBR)) +
  geom_bar()
```

**Visualising the business**

```{r}
tmap_mode('view')
tm_shape(biz_56111_sf)+
  tm_dots()
```
