---
title: "Hands-on Exercise 3"
author: "PHAM Hung Son"
date: "Sep 09, 2025"
date-modified: "last-modified"
format: 
  html:
    toc-depth: 4
    toc-expand: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Spatio-Temporal Point Patterns Analysis**

## **1. Learning Outcome**

The specific questions we would like to answer are:

-   are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally independent?

-   if the answer is NO, where and when the observed forest fire locations tend to cluster?

## **2. The data**

-   *forestfires*, a csv file provides locations of forest fire detected from the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor data. The data are downloaded from [Fire Information for Resource Management System](https://firms.modaps.eosdis.nasa.gov/download/). For the purpose of this exercise, only forest fires within Kepulauan Bangka Belitung will be used.

-   *Kepulauan_Bangka_Belitung*, an ESRI shapefile showing the sub-district (i.e. [kelurahan](https://en.wikipedia.org/wiki/Villages_of_Indonesia)) boundary of Kepulauan Bangka Belitung. The data set was downloaded from [Indonesia Geospatial](https://www.indonesia-geospasial.com/2023/05/download-shapefile-batas-administrasi.html) portal. The original data covers the whole Indonesia. For the purpose of this exercise, only sub-districts within Kepulauan Bangka Belitung are extracted.

## **3. Importing and Preparing Study Area & Forest Fire data**

```{r}
pacman::p_load(sf, terra, spatstat, sparr, tmap, tidyverse)
```

```{r}
kbb <- st_read(dsn="C:/sonphamsmu/isss626-aug2526/Hands-on_Ex/Hands-on_ex03/data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") 
```

```{r}
kbb_sf <- st_read(dsn="C:/sonphamsmu/isss626-aug2526/Hands-on_Ex/Hands-on_ex03/data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

```{r}
class(kbb_owin)
```

```{r}
fire_sf <- read_csv("C:/sonphamsmu/isss626-aug2526/Hands-on_Ex/Hands-on_ex03/data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
                       crs = 4326) %>%
  st_transform(crs = 32748)
```

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

## **4. Visualising the Fire Points**

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

```{r}
tm_shape(kbb_sf) +
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by = "Month_fac",
          ncol = 4,       # 4 columns
          nrow = 3,       # 3 rows
          free.coords = FALSE,
          drop.units = TRUE)

```

## **5. Computing STKDE by Month**

### **Extracting forest fires by month**

```{r}
fire_month <- fire_sf %>% 
  select(Month_num)
```

### **Creating ppp**

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

```{r}
summary(fire_month_ppp)
```

```{r}
any(duplicated(fire_month_ppp))
```

### **Including Owin object**

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

```{r}
plot(fire_month_owin)
```

### **Computing Spatio-temporal KDE**

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

### **Plotting the spatio-temporal KDE object**

```{r}
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

## **6. Computing STKDE by Day of Year**

### **Creating ppp object**

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

### **Including Owin object**

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

### **Computing Spatio-temporal KDE**

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

```{r}
plot(kde_yday)
```

## **7. Computing STKDE by Day of Year: Improved method**

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

### **Computing spatio-temporal KDE**

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

```{r}
plot(kde_yday)
```
