---
title: "Hands-on Exercise 2"
author: "PHAM Hung Son"
date: "Sep 03, 2025"
date-modified: "last-modified"
format: 
  html:
    toc-depth: 4
    toc-expand: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Part 1: First-order Spatial Point Patterns Analysis Methods

## **Importing and Wrangling Data**

```{r}
pacman::p_load(sf, terra, spatstat, 
               tmap, rvest, tidyverse)
```

```{r}
mpsz_sf <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
write_rds(mpsz_cl, 
          "data/mpsz_cl.rds")
```

```{r}
childcare_sf <- st_read("data/ChildCareServices.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

```{r}
tm_shape(mpsz_cl) +
  tm_polygons() +
  tm_shape(childcare_sf) +
  tm_dots()
```

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

```{r}
tmap_mode('plot')
```

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

```{r}
summary(childcare_ppp)
```

```{r}
sg_owin <- as.owin(mpsz_cl)
```

```{r}
class(sg_owin)
```

```{r}
plot(sg_owin)
```

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
childcareSG_ppp
```

## **Clark-Evan Test for Nearest Neighbour Analysis**

The test hypotheses are:

Ho = The distribution of childcare services are randomly distributed.

H1= The distribution of childcare services are not randomly distributed.

The 95% confident interval will be used.

#### Without CSR method:

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"))
```

**Result**: Since **R = 0.53532 \< 1** and **p-value \< 2.2e-16** (significant), we reject the null hypothesis.

**Conclusion**: The childcare centers are **not randomly distributed** but show **statistically significant clustering**.

#### With CSR method:

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

**Result**: R \< 1 and **p = 0.01** (significant at the 5% level.)

**Conclusion**: Again **Reject H₀**. The distribution of childcare centers is **statistically significantly clustered, not random**.

**Business Communication**

Above analysis shows that childcare centers in Singapore are not randomly scattered across the city. Instead, they are statistically clustered, with some areas having a high density of centers while other neighborhoods have few or none.

This points to potential gaps in childcare accessibility that may require policy interventions or incentives to encourage new facilities in underserved regions.

For private operators, the clustering suggests high competition in certain areas but also highlights opportunities in neighborhoods where childcare services are lacking.

## **Kernel Density Estimation Method**

```{r}
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 
```

```{r}
plot(kde_SG_diggle)
```

```{r}
summary(kde_SG_diggle)
```

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

```{r}
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp, 1000, "km")
```

```{r}
kde_childcareSG_km <- density(childcareSG_ppp_km,
                              sigma=bw.diggle,
                              edge=TRUE,
                              kernel="gaussian")
```

```{r}
plot(kde_childcareSG_km)
```

```{r}
bw.CvL(childcareSG_ppp_km)
bw.scott(childcareSG_ppp_km)
bw.ppl(childcareSG_ppp_km)
bw.diggle(childcareSG_ppp_km)
```

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp_km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

Different kernel methods:

```{r}
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

## **Fixed and Adaptive KDE**

```{r}
kde_childcareSG_fb <- density(childcareSG_ppp_km,
                              sigma=0.6, 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_childcareSG_fb)
```

```{r}
kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km, 
  method="kernel")
plot(kde_childcareSG_ab)
```

```{r}
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
```

## **Plotting cartographic quality KDE map**

```{r}
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)
```

```{r}
class(kde_childcareSG_bw_terra)
cat("\n")
kde_childcareSG_bw_terra
```

```{r}
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"
kde_childcareSG_bw_terra
```

```{r}
tm_shape(kde_childcareSG_bw_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
            title = "Density values",
            title.size = 0.7,
            text.size = 0.7,
            bg.color = "white",
            bg.alpha = 0.7,
            position = tm_pos_in(
              "right", "bottom"),
            frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```

## **First Order SPPA at the Planning Subzone Level**

```{r}
pg <- mpsz_cl %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>%
  filter(PLN_AREA_N == "JURONG WEST")
```

```{r}
plot(st_geometry(pg), main = "Punggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
```

```{r}
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

```{r}
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]
```

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

```{r}
plot(unmark(childcare_pg_ppp.km), 
  main="Punggol")
plot(unmark(childcare_tm_ppp.km), 
  main="Tampines")
plot(unmark(childcare_ck_ppp.km), 
  main="Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), 
  main="Jurong West")
```

Clark and Evans Test on Choa Chu Kang area:

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

Tampines area:

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

```{r}
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Jurong West")
```

# **Part 2: 2nd Order Spatial Point Patterns Analysis Methods**

```{r}
set.seed(1234)
```

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border") 
plot(G_CK, xlim=c(0,500))
```

#### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with G-fucntion

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

```{r}
plot(G_CK.csr)
```

**Tampines area:**

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best") 
plot(G_tm)
```

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected is p-value is smaller than alpha value of 0.001.

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
```

```{r}
plot(G_tm.csr)
```

## **Analysing Spatial Point Process Using F-Function**

**CCK area:**

```{r}
F_CK = Fest(childcare_ck_ppp) 
plot(F_CK)
```

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

Monte Carlo test with F-fucntion

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
plot(F_CK.csr)
```

**Tampines area:**

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best") 
plot(F_tm)
```

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Tampines are randomly distributed.

H1= The distribution of childcare services at Tampines are not randomly distributed.

The null hypothesis will be rejected is p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
```

```{r}
plot(F_tm.csr)
```

## **Analysing Spatial Point Process Using K-Function**

**CCK area:**

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley") 
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

**Tampines area:**

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley") 
plot(K_tm, . -r ~ r,       ylab= "K(d)-r", xlab = "d(m)",       xlim=c(0,1000))
```

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_tm.csr, . - r ~ r,       xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

## **Analysing Spatial Point Process Using L-Function**

**CCK area:**

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley") 
plot(L_ck, . -r ~ r,       ylab= "L(d)-r", xlab = "d(m)")
```

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```
