---
title: "In-class Exercise 10: Modelling Spatial Interaction"
author: "PHAM Hung Son"
date: "Nov 15, 2025"
date-modified: "last-modified"
format: 
  html:
    toc-depth: 4
    toc-expand: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## **Getting Started**

```{r}
pacman::p_load(tmap, MCI, sf, ggstatsplot, tidyverse)
```

## **Importing Data**

### **Importing geospatial data**

```{r}
stores <- st_read(dsn = "data/geospatial", layer = "STORES_TC")
```

```{r}
town <- st_read(dsn = "data/geospatial", layer = "TOWN_TC") 
```

```{r}
village <- st_read(
  dsn = "data/geospatial",
  layer = "VILLAGE_TC")
```

```{r}
tmap_mode("plot")
tm_shape(town) +
  tm_polygons() +
tm_shape(village) +
  tm_polygons() + 
tm_shape(stores) +
  tm_dots()
```

### **Importing aspatial data**

```{r}
pop2019 <- read_csv(
  "data/aspatial/pop2019.csv")

store_sales <- read_csv(
  "data/aspatial/Store_sales_data.csv")
```

```{r}
summary(store_sales)
```

## **Data Preparation**

### **Aggregating store sales**

```{r}
store_sales_aggre <- store_sales %>%
  group_by(`Store Code`) %>%
  summarise(
    "Total_Bill" = sum(Bills, na.rm = TRUE),
    "Total_Sales" = sum(Amount, na.rm = TRUE)) #na.rm to skip NA values
```

### **Relational join**

In the code chunk below, `left_join()` of **dplyr** package is used to append data from store_sales data.frame onto stores sf data.frame. At the same time a new field called *Store_size* has been created.

```{r}
stores <- stores %>%
  left_join(store_sales_aggre,
            by = join_by(
              Store_CD == `Store Code`)) %>%
  mutate(Store_size = 100)
```

## **Tidying pop2019 data.frame**

We are planning to append data in pop2019 data.frame onto village sf data.frame. Before we can do so, it is wise to check in to values in the join field are the same.

```{r}
pop2019 <- pop2019 %>%
  mutate(V_ID = str_remove_all(
    V_ID, "-"))
```

```{r}
village <- village %>%
  left_join(pop2019,
            by = join_by(
              VILLCODE == V_ID))
```

### **Check for missing or zero values**

Code chunk below is used to check in any records with missing values.

```{r}
summary(village)
```

The output print above revealed that the are five missing values in H_CNT, P_CNT, M_CNT and F_CNT fields. Hence, code chunk below is used to remove the five records with missing values from village sf data.frame.

```{r}
village <- village %>%
  filter(!is.na(P_CNT)) 
```

## **Computing Distance Matrix**

Instead of calculate the distance between the villages and stores, we will calculate the centroid of each village using the code chunk below. Then, compute the distance between the centroids of the villages and the stores.

```{r}
village_center <- st_point_on_surface(village)
distmat <- st_distance(village_center, stores) 
```

Next, the code chunk below is used to convert the distance matrix into long data.frame format

```{r}
rownames(distmat) <- village$`VILLCODE`
colnames(distmat) <- stores$`Store_CD`

distmat_long <- distmat %>%
  as.data.frame() %>%
  rownames_to_column(var = "VILLCODE") %>%
  pivot_longer(
    cols = -VILLCODE,
    names_to = "Store_CD",
    values_to = "distance"
  ) %>%
  mutate(distance = as.numeric(distance)) 
```

## **Preparing Interaction Matrix**

Code chunk below will be used to build the interaction matrix required by MCI package.

```{r}
interaction_matrix <- distmat_long %>%
  left_join(
    stores %>% 
      st_drop_geometry() %>% 
      select(Store_CD, Store_size),
    by = "Store_CD") %>%
  relocate(
    distance, .after = last_col())
```

```{r}
summary(interaction_matrix)
```

## **Computing Huff model**

```{r}
huff_share <- huff.shares(
  interaction_matrix, "VILLCODE", "Store_CD", 
  "Store_size", "distance")
```

### **Trade area analysis for a single store**

First, let us select market share data of a store by usign *Store_CD* field (i.e. TG).

```{r}
market_share <- huff_share %>%
  filter(Store_CD == "TG")
```

Next, trade areas of the selected will be derived

```{r}
Trade_area <- village %>%
  left_join(market_share)
```

Before plotting the trade area map, code chunk below is used to extract the selected store from stores sf data.frame.

```{r}
selected_store <- stores %>%
  filter(Store_CD == "TG")
```

```{r}
tm_shape(Trade_area) +
  tm_polygons(fill = "p_ij",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Patronage Probability")) +
  tm_shape(selected_store) +
    tm_dots(size = 0.3,
            fill = "red",
            col = "black") +
  tm_title("Distribution of Patronage Probability of Store TG by village") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha =0.2) 
```

### **Store sales analysis**

One of the usage of Huff model is to estimate the store sales by using village population.

Firstly, let us append P_CNT field from *village* sf.data.frame to *huff_share* data.frame by using the code chunk below.

```{r}
huff_share_pp <- huff_share %>%
  left_join(village %>% 
      st_drop_geometry() %>% 
      select(VILLCODE, P_CNT),
    by = "VILLCODE") %>%
  relocate(
    distance, .after = last_col())
```

Next, `shares.total()` of **MCI** package will be used to calculate the total sales by using the code chunk below.

```{r}
huff_total <- shares.total(
  huff_share_pp, "VILLCODE", "Store_CD",
  "p_ij", "P_CNT")
```

Code chunk below is used to append the estimate store sales (Ej) onto stores sf data.frame.

```{r}
store_sales_est <- stores %>%
  left_join(huff_total,
            by = join_by(
              "Store_CD" == "suppliers_single"))
```

Now, we can visualise how well the estimate sales as compare to the actual sales

```{r}
store_sales_nongeom <- store_sales_est %>% st_drop_geometry()

ggscatterstats(
  store_sales_nongeom,
  x = sum_E_j,
  y = Total_Sales)
```

```{r}
huff_share_wider <- huff_share %>%
  select(VILLCODE, Store_CD, p_ij) %>%
    pivot_wider(names_from = Store_CD, 
                values_from = p_ij)
```
