---
title: "In-class Exercise 9: Modelling Geographical Accessibility"
author: "PHAM Hung Son"
date: "Nov 08, 2025"
date-modified: "last-modified"
format: 
  html:
    toc-depth: 4
    toc-expand: true
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## **Getting Started**

```{r}
pacman::p_load(tmap, SpatialAcc, sf, 
               ggstatsplot, reshape2,
               tidyverse)
```

## **Geospatial Data Wrangling**

### **Importing geospatial data**

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

### **Updating CRS information**

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

### **Cleaning and updating attribute fields of the geospatial data**

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

## **Apsaital Data Handling and Wrangling**

### **Importing Distance Matrix**

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### **Tidying distance matrix**

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

```{r}
distmat_km <- as.matrix(distmat/1000)
```

## **Modelling and Visualising Accessibility using Hansen Method**

### **Computing Hansen’s accessibility**

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 0.5, 
                            family = "Hansen"))

colnames(acc_Hansen) <- "accHansen"
acc_Hansen <- as_tibble(acc_Hansen)
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

```{r}
acc_Hansen1 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 1, 
                            family = "Hansen"))

colnames(acc_Hansen1) <- "accHansen1"
acc_Hansen1 <- as_tibble(acc_Hansen1)
hexagon_Hansen1 <- bind_cols(hexagons, acc_Hansen1)
```

```{r}
acc_Hansen15 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 1.5, 
                            family = "Hansen"))

colnames(acc_Hansen15) <- "accHansen15"
acc_Hansen15 <- as_tibble(acc_Hansen15)
hexagon_Hansen15 <- bind_cols(hexagons, acc_Hansen15)
```

```{r}
acc_Hansen20 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2.0, 
                            family = "Hansen"))

colnames(acc_Hansen20) <- "accHansen20"
acc_Hansen20 <- as_tibble(acc_Hansen20)
hexagon_Hansen20 <- bind_cols(hexagons, acc_Hansen20)
```

```{r}
acc_Hansen25 <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2.5, 
                            family = "Hansen"))

colnames(acc_Hansen25) <- "accHansen25"
acc_Hansen25 <- as_tibble(acc_Hansen25)
hexagon_Hansen25 <- bind_cols(hexagons, acc_Hansen25)
```

### **Visualising Hansen’s accessibility**

**Extracting map extend**

```{r}
mapex <- st_bbox(hexagons)
```

```{r}
tmap_mode("plot")
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Hansen 0.5") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)

tm_shape(hexagon_Hansen1,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen1", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Hansen 1.0") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)

tm_shape(hexagon_Hansen15,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen15", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Hansen 1.5") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)

tm_shape(hexagon_Hansen20,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen20", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Hansen 2.0") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)

tm_shape(hexagon_Hansen25,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen25", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Hansen 2.5") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```
